---
title: "Million Song Dataset"
author: "Astri Cahyani"
execute:
  echo: false
---


```{r}
#| message: false
library(tidyverse)
```

::: {.insights-box}
### Echo Nest Music Metrics
*   **Artist Familiarity:** An algorithmic estimation (scaled 0 to 1) of how well-known an artist is to listeners.
*   **Artist Hotttnesss:** A measurement (scaled 0 to 1) of the artist's current popularity or "buzz" at the time of data collection.
:::

```{r}
#| message: false
# Read and print the music dataset
music <- read_csv("data/music.csv", show_col_types = FALSE)
music %>% 
  select(starts_with("artist")) %>% 
  glimpse()
```
```{r}
#| message: false
# Filter out songs with a release year of 0
music_filtered <- music %>% 
  filter(song.year != 0)

# Plot the distribution of release years
ggplot(music_filtered, aes(x = song.year)) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Distribution of Song Release Years",
    subtitle = paste("Number of songs:", nrow(music_filtered)),
    x = "Year",
    y = "Count"
  )
```
```{r}
# Calculate summary statistics (minimum, maximum, median, and quartiles) for selected columns
music %>%
  select(artist.familiarity, artist.hotttnesss, song.year, song.tempo) %>%
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value") %>%
  group_by(variable) %>%
  summarize(
    min = min(value, na.rm = TRUE),
    q25 = quantile(value, 0.25, na.rm = TRUE),
    median = median(value, na.rm = TRUE),
    q75 = quantile(value, 0.75, na.rm = TRUE),
    max = max(value, na.rm = TRUE)
  )
```
```{r}
# Count placeholder values (0) in the specified columns
music %>%
  select(artist.location, release.name, song.title, song.year, artist.familiarity, artist.hotttnesss) %>%
  pivot_longer(cols = everything(), names_to = "column", values_to = "value") %>%
  group_by(column) %>%
  summarize(
    placeholders = sum(value == 0, na.rm = TRUE)
  )
```

::: {.insights-box}
### Missing Location Coordinates
A large portion of artists lack actual geographical coordinates, with their latitude and longitude set to a default placeholder of `(0,0)`.
:::

```{r}
# Count unique artists with usable vs placeholder coordinates
music %>%
  distinct(artist.id, artist.latitude, artist.longitude) %>%
  mutate(
    coordinate_type = if_else(artist.latitude == 0 & artist.longitude == 0, "placeholder", "usable")
  ) %>%
  count(coordinate_type)
```

::: {.insights-box}
### Geographic Coverage Note
Because **more than 60%** of the artists in the dataset lack valid coordinates, the world map below only displays the subset of artists with usable location data.
:::

```{r}
#| message: false
library(maps)

# Filter for songs with usable coordinates and non-zero years
music_usable <- music %>%
  filter(!(artist.latitude == 0 & artist.longitude == 0)) %>%
  filter(song.year != 0)

# Get world map polygons
world_map <- map_data("world")

# Plot world map with artist locations colored by song year
ggplot() +
  geom_polygon(data = world_map, aes(x = long, y = lat, group = group), fill = "#E0E0E0", color = "white") +
  geom_point(data = music_usable, aes(x = artist.longitude, y = artist.latitude, color = song.year), alpha = 0.6, size = 1.2) +
  scale_color_viridis_c(option = "viridis") +
  coord_quickmap() +
  labs(
    title = "Geographical Distribution of Artists by Song Release Year",
    subtitle = "Color represents song release year (excludes placeholder coordinates and year = 0)",
    x = "Longitude",
    y = "Latitude",
    color = "Release Year"
  ) +
  theme_minimal()
```





