---
title: "Analyzing Music Data"
author: "Astri Cahyani"
execute:
  echo: false
---


```{r}
#| message: false
library(tidyverse)
```

```{r}
billboard %>% 
  select(artist, track, date.entered, wk1, wk2, wk3, wk4)
```
```{r}
billboard %>% 
  summarize(
    min = min(date.entered),
    max = max(date.entered)
  )
```
```{r}
#| message: false
ggplot(billboard, aes(x = wk1)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  labs(
    title = "Histogram of Week 1 Rankings",
    subtitle = paste("Number of values:", sum(!is.na(billboard$wk1))),
    x = "Week 1 Rank",
    y = "Count"
  )
```
```{r}
#| message: false
ggplot(billboard, aes(x = wk6)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  labs(
    title = "Histogram of Week 6 Rankings",
    subtitle = paste("Number of values:", sum(!is.na(billboard$wk6))),
    x = "Week 6 Rank",
    y = "Count"
  )
```
```{r}
billboard %>%
  select(wk1, wk4, wk10, wk20, wk40, wk76) %>%
  pivot_longer(cols = everything(), names_to = "week", values_to = "rank") %>%
  mutate(week = factor(week, levels = c("wk1", "wk4", "wk10", "wk20", "wk40", "wk76"))) %>%
  group_by(week) %>%
  summarize(
    present = sum(!is.na(rank)),
    missing = sum(is.na(rank))
  )
```
```{r}
billboard %>%
  pivot_longer(
    cols = starts_with("wk"),
    names_to = "week",
    values_to = "rank"
  ) %>%
  mutate(week_num = as.integer(str_remove(week, "wk"))) %>%
  group_by(artist, track) %>%
  summarize(
    reentered = any(is.na(rank) & week_num < max(week_num[!is.na(rank)])),
    .groups = "drop"
  ) %>%
  count(reentered)
```
```{r}
# Count of songs by category of rank change between week 1 and week 6
billboard %>%
  mutate(
    change_category = case_when(
      is.na(wk6) ~ "missing by week 6",
      wk6 < wk1  ~ "improved",
      wk6 == wk1 ~ "stayed the same",
      wk6 > wk1  ~ "got worse"
    )
  ) %>%
  count(change_category)

# Median rank change (wk6 - wk1) for songs present in both weeks
billboard %>%
  filter(!is.na(wk1), !is.na(wk6)) %>%
  summarize(
    median_change = median(wk6 - wk1)
  )
```
```{r}
#| cache: true
# Reshape to long format and plot each song's ranking over time
billboard_long <- billboard %>%
  pivot_longer(
    cols = starts_with("wk"),
    names_to = "week",
    values_to = "rank",
    values_drop_na = TRUE
  ) %>%
  mutate(week_num = as.integer(str_remove(week, "wk")))

ggplot(billboard_long, aes(x = week_num, y = rank, group = interaction(artist, track))) +
  geom_line(alpha = 0.2, color = "steelblue") +
  scale_y_reverse() +
  labs(
    title = "Billboard Song Rankings Over Time",
    x = "Week on Chart",
    y = "Rank (Reversed)"
  )
```
```{r}
# Create the song-level summary
song_summary <- billboard %>%
  pivot_longer(
    cols = starts_with("wk"),
    names_to = "week",
    values_to = "rank",
    values_drop_na = TRUE
  ) %>%
  mutate(week_num = as.integer(str_remove(week, "wk"))) %>%
  group_by(artist, track) %>%
  summarize(
    first_rank = rank[week_num == 1],
    best_rank = min(rank),
    first_week_at_best = min(week_num[rank == best_rank]),
    total_weeks = n(),
    .groups = "drop"
  )

# Display a preview of the song summary (first few rows)
head(song_summary)

# Identify the three notable top-10 songs
fastest_no1 <- song_summary %>%
  filter(best_rank == 1) %>%
  slice_min(first_week_at_best, n = 1)

slowest_no1 <- song_summary %>%
  filter(best_rank == 1) %>%
  slice_max(first_week_at_best, n = 1)

longest_run_top10 <- song_summary %>%
  filter(best_rank <= 10) %>%
  slice_max(total_weeks, n = 1)

# Print the three notable songs in a summary table
bind_rows(
  mutate(fastest_no1, note = "Fastest to #1"),
  mutate(slowest_no1, note = "Slowest to #1"),
  mutate(longest_run_top10, note = "Longest run top-10")
) %>%
  select(note, artist, track, first_rank, best_rank, first_week_at_best, total_weeks)
```

::: {.insights-box}
### Year 2000 Billboard Chart Insights
An analysis of the **317 tracks** entering the Billboard Hot 100 in the year 2000 reveals three standout records:

*   **Fastest Climb to #1:** Madonna’s *"Music"* reached the top spot in just **6 weeks** (entering the chart at #41).
*   **Slowest Climb to #1:** Lonestar’s *"Amazed"* took **40 weeks** of climbing to reach #1 (entering at #81).
*   **Longest Run on Chart:** Creed’s *"Higher"* set the endurance record, spending **57 weeks** on the chart (peaking at #7).
:::

```{r}
# Filter to only the songs that reached the top 10 at some point
top10_songs <- billboard_long %>%
  group_by(artist, track) %>%
  filter(min(rank) <= 10) %>%
  ungroup() %>%
  mutate(
    highlight = case_when(
      artist == "Madonna" & track == "Music" ~ "Madonna - Music (Fastest to #1)",
      artist == "Lonestar" & track == "Amazed" ~ "Lonestar - Amazed (Slowest to #1)",
      artist == "Creed" & track == "Higher" ~ "Creed - Higher (Longest Top-10 Run)",
      TRUE ~ "Other Top-10 Songs"
    )
  )

# Plot the top-10 songs with highlighting
ggplot(top10_songs, aes(x = week_num, y = rank, group = interaction(artist, track), color = highlight)) +
  # Render the gray lines first
  geom_line(data = filter(top10_songs, highlight == "Other Top-10 Songs"), alpha = 0.4) +
  # Render the colored highlight lines on top
  geom_line(data = filter(top10_songs, highlight != "Other Top-10 Songs"), linewidth = 1.2) +
  scale_y_reverse() +
  scale_color_manual(
    values = c(
      "Madonna - Music (Fastest to #1)" = "#ff7ebb",
      "Lonestar - Amazed (Slowest to #1)" = "#70d6ff",
      "Creed - Higher (Longest Top-10 Run)" = "#c77dff",
      "Other Top-10 Songs" = "#d1d5db"
    ),
    breaks = c(
      "Madonna - Music (Fastest to #1)",
      "Lonestar - Amazed (Slowest to #1)",
      "Creed - Higher (Longest Top-10 Run)"
    )
  ) +
  labs(
    title = "Top-10 Songs on the Billboard Chart",
    subtitle = "Highlighting three notable songs",
    x = "Week on Chart",
    y = "Rank (Reversed)",
    color = "Notable Songs"
  )
```
