---
title: "NBA Scorigami"
author: "Astri Cahyani"
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| message: false
#| warning: false
library(tidyverse)
```

```{r}
#| message: false
#| warning: false
games <- read_csv("data/Games.csv")
```

::: {.insights-box}
### What is Scorigami?
A **Scorigami** is a final score combination that has occurred for the first time in league history. Out of **73,277 NBA games** in this dataset, teams have registered **5,119 unique score combinations**.

*   **Home-Court Advantage:** Home teams won **61.57%** of all games, visible in the higher density of scores below the diagonal line ($Home\ PTS > Visitor\ PTS$).
*   **Most Common Score:** **105 - 99** (a home-team victory) is the most frequent score, occurring **100 times**.
*   **Extreme Outlier:** The highest-scoring game occurred on Dec 13, 1983, when the Detroit Pistons defeated the Denver Nuggets **186 - 184** (370 total points) in triple overtime.
:::

::: {.scorigami-wrapper}
```{r}
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 10
library(plotly)

# Filter out empty or incomplete games
games_filtered <- games %>%
  filter(homeScore > 0, awayScore > 0)

# Compute counts and details for tooltip
score_counts_interactive <- games_filtered %>%
  group_by(homeScore, awayScore) %>%
  summarize(
    n = n(),
    first_date = format(min(gameDate), "%b %d, %Y"),
    last_date = format(max(gameDate), "%b %d, %Y"),
    sample_teams = paste0(first(paste(hometeamCity, hometeamName)), " vs ", first(paste(awayteamCity, awayteamName))),
    .groups = "drop"
  ) %>%
  mutate(
    tooltip_text = paste0(
      "Score: ", homeScore, " - ", awayScore, "\n",
      "Occurrences: ", n, "\n",
      "First Achieved: ", first_date, "\n",
      "Most Recent: ", last_date, "\n",
      "Example: ", sample_teams
    )
  )

# Plot
p <- ggplot(score_counts_interactive, aes(x = homeScore, y = awayScore)) +
  geom_tile(aes(fill = n, text = tooltip_text), width = 0.95, height = 0.95) +
  geom_abline(intercept = 0, slope = 1, color = "white", linewidth = 1.2) +
  scale_fill_viridis_c(
    option = "inferno", 
    trans = "log10", 
    name = "Count",
    breaks = c(1, 10, 100),
    labels = c("1", "10", "100+")
  ) +
  scale_x_continuous(
    breaks = seq(20, 180, by = 20), 
    minor_breaks = seq(10, 200, by = 2),
    limits = c(15, 195),
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    breaks = seq(20, 180, by = 20), 
    minor_breaks = seq(10, 200, by = 2),
    limits = c(15, 195),
    expand = c(0, 0)
  ) +
  labs(
    title = "NBA Scorigami",
    x = "Home PTS",
    y = "Visitor PTS"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 22, color = "#0B1D3A", hjust = 0, margin = margin(b = 6)),
    axis.title = element_text(face = "bold", size = 14, color = "grey20"),
    axis.text = element_text(size = 12, color = "grey30"),
    panel.background = element_rect(fill = "#EAEAEA", color = NA),
    panel.grid.major = element_line(color = "white", linewidth = 0.4),
    panel.grid.minor = element_line(color = "white", linewidth = 0.15),
    plot.background = element_rect(fill = "white", color = NA),
    aspect.ratio = 1,
    legend.position = "right"
  )

# Render plotly widget
ggplotly(p, tooltip = "text") %>%
  layout(
    xaxis = list(scaleanchor = "y", scaleratio = 1),
    margin = list(t = 50)
  )
```
:::

## Scoring Trends and Scorigami Rates

How does the pace and scoring of the game affect how often we see a new final score? The charts below illustrate the relationship between the league's average combined points per game and the rate of new Scorigami discoveries from 1946 to 2026.

::: {.scorigami-wrapper}
```{r}
#| message: false
#| warning: false
#| fig-width: 9
#| fig-height: 7

# Calculate yearly average combined points
yearly_scoring <- games %>%
  filter(homeScore > 0, awayScore > 0) %>%
  mutate(year = year(gameDate)) %>%
  group_by(year) %>%
  summarize(
    avg_total_pts = mean(homeScore + awayScore),
    .groups = "drop"
  )

# Calculate new scorigamis per year
new_scorigamis <- games %>%
  filter(homeScore > 0, awayScore > 0) %>%
  arrange(gameDate) %>%
  distinct(homeScore, awayScore, .keep_all = TRUE) %>%
  mutate(year = year(gameDate)) %>%
  group_by(year) %>%
  summarize(new_scores_count = n(), .groups = "drop")

# Merge
merged_stats <- full_join(yearly_scoring, new_scorigamis, by = "year") %>%
  replace_na(list(new_scores_count = 0)) %>%
  pivot_longer(
    cols = c(avg_total_pts, new_scores_count),
    names_to = "metric",
    values_to = "value"
  ) %>%
  mutate(
    metric_label = factor(
      metric,
      levels = c("avg_total_pts", "new_scores_count"),
      labels = c("Average Combined Points per Game", "New Scorigami Combinations Discovered")
    )
  )

# Plot
ggplot(merged_stats, aes(x = year, y = value, fill = metric_label, color = metric_label)) +
  geom_area(alpha = 0.15, position = "identity") +
  geom_line(linewidth = 1.2) +
  facet_wrap(~metric_label, scales = "free_y", ncol = 1) +
  scale_x_continuous(breaks = seq(1946, 2026, by = 10)) +
  scale_fill_manual(values = c("Average Combined Points per Game" = "#70d6ff", "New Scorigami Combinations Discovered" = "#ff7ebb")) +
  scale_color_manual(values = c("Average Combined Points per Game" = "#3b82f6", "New Scorigami Combinations Discovered" = "#db2777")) +
  labs(
    title = "NBA Scoring Pace vs. Scorigami Rates (1946 - 2026)",
    subtitle = "Spikes in game pace and scoring trigger waves of newly discovered score combinations",
    x = "Year",
    y = NULL,
    caption = "Data source: kaggle.com"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16, color = "#0B1D3A"),
    plot.subtitle = element_text(size = 11, color = "grey30"),
    strip.text = element_text(face = "bold", size = 11, color = "grey20", hjust = 0),
    panel.background = element_rect(fill = "#F9FAFB", color = NA),
    panel.grid.major = element_line(color = "white", linewidth = 0.5),
    panel.grid.minor = element_line(color = "white", linewidth = 0.2),
    plot.background = element_rect(fill = "white", color = NA),
    legend.position = "none"
  )
```
:::

::: {.insights-box}
### Pace and Scorigami Insights
*   **The Shot Clock Jump (1954):** The introduction of the 24-second shot clock in 1954 caused average combined scores to surge from **159 to over 190 points** in just a few seasons, causing a massive wave of new final scores (peaking at **178 new combinations** in a single year).
*   **The Scorigami Drought (1970s–2000s):** As the league's pace stabilized and the grid of common scores filled up, the discovery of new Scorigamis dropped below **30 per year** in the early 2000s, coinciding with a low-scoring defensive era (averaging around **190 combined points**).
*   **The Modern Resurgence (2010s–2020s):** High-volume three-point shooting and modern transition play have pushed the average combined points up to **225+**. This modern offensive explosion has unlocked previously untouched high-scoring regions, causing new Scorigami rates to double to **40+ per year**.
:::

## Deep Dive: NBA Scoring Analytics

To better understand the mechanics of NBA Scorigami, let's explore three key visualizations detailing score frequencies, decade-by-decade probability patterns, and score margins.

### 1. The Most Frequent Scores
While Scorigamis represent the *first* time a score occurs, most NBA games cluster around a few very common score combinations.

::: {.scorigami-wrapper}
```{r}
#| message: false
#| warning: false
#| fig-width: 9
#| fig-height: 5

top_scores_interactive <- games_filtered %>%
  count(homeScore, awayScore, sort = TRUE) %>%
  head(15) %>%
  mutate(
    score_label = paste(homeScore, "-", awayScore),
    score_label = reorder(score_label, n),
    tooltip_text = paste0("Score: ", homeScore, " - ", awayScore, "\nTotal Occurrences: ", n)
  )

p_top <- ggplot(top_scores_interactive, aes(y = score_label, x = n, fill = n, text = tooltip_text)) +
  geom_bar(stat = "identity", width = 0.7) +
  scale_fill_gradient(low = "#ff7ebb", high = "#c77dff") +
  labs(
    title = "Top 15 Most Frequent Final Scores",
    x = "Total Games Played",
    y = "Final Score (Home - Away)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 15, color = "#0B1D3A"),
    panel.background = element_rect(fill = "#F9FAFB", color = NA),
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(color = "white", linewidth = 0.5)
  )

ggplotly(p_top, tooltip = "text") %>%
  layout(margin = list(l = 80))
```
:::

### 2. Scorigami Probability by Decade
How hard is it to find a Scorigami? In the early days of the league, nearly every game resulted in a new score. By the 2010s, it became extremely rare. However, the 2020s are experiencing a sudden resurgence.

::: {.scorigami-wrapper}
```{r}
#| message: false
#| warning: false
#| fig-width: 9
#| fig-height: 5

scorigami_pct <- games_filtered %>%
  arrange(gameDate) %>%
  mutate(
    score_id = paste(homeScore, awayScore),
    is_scorigami = !duplicated(score_id),
    decade = paste0(floor(year(gameDate) / 10) * 10, "s")
  ) %>%
  group_by(decade) %>%
  summarize(
    total_games = n(),
    scorigamis = sum(is_scorigami),
    pct_scorigami = scorigamis / total_games * 100,
    .groups = "drop"
  ) %>%
  mutate(
    tooltip_text = paste0(
      "Decade: ", decade, "\n",
      "Total Games: ", scales::comma(total_games), "\n",
      "New Scorigamis: ", scorigamis, "\n",
      "Probability: ", sprintf("%.2f%%", pct_scorigami)
    )
  )

p_prob <- ggplot(scorigami_pct, aes(x = decade, y = pct_scorigami, fill = pct_scorigami, text = tooltip_text)) +
  geom_bar(stat = "identity", width = 0.6) +
  scale_fill_gradient(low = "#c77dff", high = "#ff7ebb") +
  labs(
    title = "Percentage of Games Resulting in a Scorigami",
    x = "Decade",
    y = "% of Games"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 15, color = "#0B1D3A"),
    panel.background = element_rect(fill = "#F9FAFB", color = NA),
    panel.grid.major.x = element_blank()
  )

ggplotly(p_prob, tooltip = "text")
```
:::

### 3. Distribution of Victory Margins
Due to home-court advantage, victory margins are heavily skewed in favor of the home team, with home wins outpacing visitor wins **61.57% to 38.43%**.

::: {.scorigami-wrapper}
```{r}
#| message: false
#| warning: false
#| fig-width: 9
#| fig-height: 5

margin_counts <- games_filtered %>%
  mutate(
    margin = homeScore - awayScore,
    winner = if_else(margin > 0, "Home Win", "Away Win")
  ) %>%
  count(margin, winner) %>%
  mutate(
    tooltip_text = paste0(
      "Victory Margin: ", abs(margin), " pts (", winner, ")\n",
      "Total Games: ", n
    )
  )

p_margin <- ggplot(margin_counts, aes(x = margin, y = n, fill = winner, text = tooltip_text)) +
  geom_bar(stat = "identity", width = 0.9, alpha = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40", linewidth = 1) +
  scale_fill_manual(values = c("Home Win" = "#70d6ff", "Away Win" = "#ff7ebb")) +
  labs(
    title = "NBA Margin of Victory Distribution",
    x = "Margin (Home PTS - Visitor PTS)",
    y = "Number of Games",
    fill = "Winner"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 15, color = "#0B1D3A"),
    panel.background = element_rect(fill = "#F9FAFB", color = NA),
    legend.position = "none"
  )

ggplotly(p_margin, tooltip = "text")
```
:::

::: {.insights-box}
### Deep Dive Insights
*   **The 6-Point Sweet Spot:** The most common final score in NBA history is **105 - 99** (and its close neighbor **108 - 101**), both representing a 6-point victory margin.
*   **Decade U-Turn:** The probability of a Scorigami dropped continuously for 70 years, hitting a low of **1.65%** in the 2010s. However, the offensive explosion of the 2020s has reversed this trend, nearly doubling the probability to **3.12%**.
*   **The Home Shift:** The distribution of victory margins is centered at a mean of **+3.54 points** in favor of the home team, indicating the strength of home-court advantage in final game outcomes.
:::

## Interactive Scorigami Explorer

### Features
*   **Filter by Team**: Select a specific NBA franchise to view only their final score combinations.
*   **Filter by Game Type**: Isolate regular season matches, Play-in Tournaments, or NBA Finals.
*   **Time-based Selection**: Use the slider to filter scores within specific decades or seasons.
*   **Hover Tooltips**: Inspect individual scores to see occurrences, first/last dates achieved, and a sample game matchup.

### How to Run Locally

To launch the dashboard locally in your R console by running:

```r
shiny::runApp("app.R")
```



